---
title: "ESM206 lab5"
author: "Shuhan Song"
date: "October 29, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
#### Lab 5 objectives

- learn to parse dates with 'lubridate'
- get counts of observations with count()
- Uncount() rows
- One and two-sample t-test
- Create a heatmap with geom_tile()

```{r, include=FALSE}
# attach packages
library(tidyverse)
library(here)
library(janitor)

# read in data
lobster_abundance <- read_csv(here::here("data", "lobster_abundance.csv"),
                              na = "-99999") %>% 
  janitor::clean_names()
# count: combined multiple observations in frequency format, which should be split into each individual oobservation
```

use tidyr::uncount() to convert out lobster data from frequency format to case format
```{r}
lobster_tidy <- lobster_abundance %>% 
  tidyr::uncount(lobster_count) # repeat each row "count" times, removed count=0 rows, can use fcn populate to find the eliminated rows
```

#### Exploratory data visualization

We're only going to consider "site" as our variable of interest
```{r}
ggplot(lobster_tidy, aes(x = site, y = size_mm)) +
  geom_jitter(aes(color = site), 
              alpha = 0.5, 
              width = 0.2)
#histogram
ggplot(lobster_tidy, aes(x = size_mm)) +
  geom_histogram(aes(fill = site)) +
  facet_wrap(~site, scales = "free") # scale fit the data in each group. only use when not comparing across groups

# QQ plot
ggplot(lobster_tidy, aes(sample = size_mm)) +
  geom_qq()+
  facet_wrap(~site)

# don't let outliers influence decisions on whether it's normal distribution or not. 
# the slop is not meaningful. only LINEAR relatinship matters. linear means normal distribution

```


#### Convert the "data" column to class "Date"

We'll use the 'lubridate' packages to convert to data format, and then to help us easily parse month & years

```{r}
lobster_date <- lobster_tidy %>% 
  mutate(
    date_new = lubridate::mdy(date) #mdy - month day year, can also use other combination
  )
```


Now let's parse year and month using lubridate::month() and lubridate::year()

```{r}
lobster_parse_date <- lobster_date %>% 
  mutate(
    obs_month = lubridate::month(date_new, label = TRUE), # default is to pull the month as NUMBER!! # label = TRUE will replace the month number with abbreviation and automatically recognized it as an ordered factor
    obs_year = lubridate::year(date_new)
  )
levels(lobster_parse_date$obs_month) # even just two month included, recognized all levels
```

Now let's find counts of observed lobsters based on different hierarchical groupings:

count lobsters by year and month
```{r}
# count by month
lobster_ym <- lobster_parse_date %>% 
  dplyr::count(obs_year, obs_month) # first group by year, then grouped by month

lobster_ym

# count by year
lobster_y <- lobster_parse_date %>% 
  dplyr::count(obs_year)

lobster_y

# count by site
lobster_site <- lobster_parse_date %>% 
  dplyr::count(site)
lobster_site

# count by site and transect
```
note: make the data in tidy format so that users can analyze the data as they prefer

If we want to create a summary table that contains statistics other than counts by group. It's easier to use group_by() + n()
```{r}
lobster_summary <- lobster_parse_date %>% 
  group_by(site) %>% 
  summarize(
    lobster_number = n(), # find counts
    mean_size = mean(size_mm, na.rm = TRUE),
    sd_size = sd(size_mm, na.rm = TRUE)
  )
lobster_summary
```

#### Find confidence intervals

Use t.test() to find CI (for one sample) and perform t-tests to compare means of two samples (this will cover conceptually in lectures week 6)
```{r}
ivee_lobster <- lobster_tidy %>% 
  filter(site == "IVEE") %>% 
  pull(size_mm) # pull() create a vector that contains all size_mm

t.test(ivee_lobster) # use for a vector containing many numbers
```
95% CI: 
If take many samples of the population, expect 95% of the samples contain the true population mean


#### Two-sample t-test to compare means

We want to ask: is there a significant difference in lobster lengths at Naples and Mohawk reefs?

We've done our necessary exploratory analyses (normal, sample size) to determine that a 2-sample t-test for means comparison is appropriate. 

* give t-test the entire samples, not the samply means
```{r}
napl_sample <- lobster_tidy %>% 
  filter(site == "NAPL") %>% 
  pull(size_mm)

mohk_sample <- lobster_tidy %>% 
  filter(site == "MOHK") %>% 
  pull(size_mm)

mn_ttest <- t.test(mohk_sample, napl_sample) # in either direction, are the two data significantly, different
# Welch: the two samples do not have the same variance

# 95% CI: true sample differences between individuals in each group
```

There is a significant difference in lobster lengths between Naples and Mohawk Reef (t(df) = statistic, p < 0.001, alpha = 0.05). 
- p-value is not the golden key. 
- data size, structure, distribution, etc. are all important


```{r}
lobster_mn <- lobster_tidy %>% 
  filter(site %in% c("NAPL", "MOHK"))

mn_ttest2 <- t.test(size_mm ~ site, data = lobster_mn) # t test on size_mm between two levels in "site" in lobster_mn
mn_ttest2
```



There is a sig diff (t(1850.8) = -19.849, p = 2.2e-16). copy and paste the values is not reproducible. 

Instead:

Here is the p-value: `r mn_ttest2$p.value`   
Here is the t-statistic: `r mn_ttest2$statistic`


#### Heat map

```{r}
lobster_ys <- lobster_parse_date %>% 
  count(obs_year, site)


ggplot(data = lobster_ys, aes(x = obs_year, y = site)) +
  geom_tile(aes(fill = n))

```




